---
title: "Achtergrondwaarden"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Achtergrondwaarden}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

To prevent soil pollution in the Netherlands, national legislation is
in place. In this legislation for many substances, like heavy metals,
organic pollatants, etc., several reference values are defined, including
the so called 'background values', or 'achtergrondwaarden' in Dutch. These
background values are soil concentrations and used to assess the quality
of the soils. From a policy point of view, soils with concentrations
below the background value are considered clean.

The background values are broadly based on the 95-percentile of measured
soil concentrations from a representative set of soil samples from a
defined area. While nationwide background values exists, local
authorities can determine their own background values, if needed. For
example in the case of specific soils or circumstances.

The choice for the 95-percentile is based on the probability that if one
takes a sample from a clean soil, that only 5%, i.e. 1 in 20 cases,
should fail the test.

Conceptually, the calculation of a 95-percentile, a rank order statistic,
is relatively simple. However the implementation can be cumbersome
considering that datasets ussually have less than 100 samples, 
and do contain outlying and censored values.

The purpose of this package is to provide several methods to calculate
a rank order statistic, often the 95 percentile, but other percentiles
are also possible. Besides this 95-percentile, confidence intervals of
this percentile are also calculated.


# Limits of quantification (LOQ)

It is common that datasets with chemical soil analyses contain values
below a limit of quantification (LOQ). This concentrations are usually
indicated with a 'less than' sign, e.g. '<1 mg/kg'. Such values can
not be considerd as numerical values, due to the '<', while numerical
values are needed for the data analyses.

One way to cope with these values is to transform them to negative
concentrations. In theory, a measured soil concentration can not be
negative, so they don't exists in the database (and if they do, you
should complain to the laboratory). The following code replaces these
'<' signs:

```{r}
library(achtergrondwaarden)
soildata <- c("<1","2","3")
x <- as.numeric(replacelt(soildata))
x
```

These negative values can then easily imputed to non-negative values
using a impution factor. The following code replaces the value
with half the LOQ value. 

```{r}

x.imp <- replaceNegative(x,replaceval=0.5)
x.imp

```

Other values which can be used are for example 0, or 0.7 in case of
log-normal distributed values. The default value is to replace the LOQ
value with NA.

# Outliers

In case of a sample size of 50, including 5 values as outliers, the
95-percentile will be based on these outlying values. It is often
unclear if an outlier is the result of a sampling or analytical
artefact or a true outlier based on variance in soil concentrations.
The choice of how to deal with outliers is therefore difficult.

In case one wants to remove the outliers from the data, several
methods are common. The achtergrondwaarden package contains a function
to remove outliers based on the distance of the median. The measure of
this distance is the median absolute deviation (MAD).

The following code removes outliers which are larger than the median
plus b=3 times the MAD

```{r}

soildata <- c(1,2,3,2,1,10)
x <- rmoutlier(soildata,b=3)
x

```

# Estimating distribution parameters

To estimate the 95-percentile and it's confidence intervals one can
use a theoretical approach. Considering a dataset of n=100, the 95
percentile is equal to the 95th value of the ordered data. The
theoretical confidence intervals can then be calculated as follows:

```{r}
library(binom)
n <- 100
ci <- binom.confint(x=0.95*n,n,methods="exact")
ci
```

given the above example, the 95 percentile and confidence intervals of
a dataset x can be estimated

```{r}
x <- rnorm(n)
quantile(x,probs=c(ci$lower,.95,ci$upper))
```

## Bootstrapping

Another way to estimate percitiles, like the 95-percentile, is using
bootstrapping. The achtergrondwaarde library uses the boot package to
perform the bootstrapping. One can estimate a 95-percentile using the
following code:

```{r}
data(meuse,package="sp")
p <- pbci(meuse$zinc,p=.95)
p

```

## Distribution fitting

A third way to estimate percentiles and confidence intervals is by
fitting a distribution model and then estimate the interval using the
distribution model parameters. For this aproach the package
fitdistrplus is used. The logfitci function from this package is a
wrapper around the fitdist and bootdist functions from fitdistrplus,
using a logarithmic model. To calculate the estimates the following 
code can be used:

```{r}

p <- logfitci(x=meuse$zinc,p=.95)
p


```

## Distribution fitting and LOQs

When the data contains many censored values, fitting the distribution can
lean to a non-optimal fit. In the fitdistrplus package the fitdistcens
function is available. This function fits a distribution based on
data including censored value.   The logfitcicens function of the
achtergrondwaarde package is a wrapper around the fitdistcens and
bootdistcens functions from fitdistrplus package.

If we assume that the limit of quantification of the zinc data lies at
the 20th-percentile and replace the values below this percentile with
the percentile itself (using negative numbers), othen one can estimate the
95-percentile and intervals as follows.

```{r}


x <- meuse$zinc
loq <- quantile(x,p=.2)
x <- ifelse(x>loq,x,-1*loq)
p <- logfitcicens(x)
p$f
p$result


```

# all together now

The qestimates function estimates quantiles based on bootstrapping and
both fitting functions mentioned above.  This qestimate function calls
respectively the pbci, logfitci and logfitcicens functions and returns
the result as a data.frame.  The function tries hard to not throw an
error when the estimation fails, in that case the estimates will be NA.

Environmental research datasets often contain many parameters.
Using a for loop or apply function the data can be gathered in one
single data.frame for further analyses. If parameters do not contain
values or have a variance of zero, resulting in an error in underlying
functions, this function returns NA values. So it can be part of an
automated procedure.

On single call to qestimate looks like this:

```{r}

p <- qestimates(meuse$zinc,p=.95)
p$result


```



